---
title: "Archival tags"
author: "James T. Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Archival tags}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{dsem}
---

```{r, include = FALSE}
EVAL <- TRUE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = EVAL,
  purl = EVAL
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\walk)', force=TRUE )
# Build and PDF
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\walk)'); devtools::build_rmd("vignettes/archival_tags.Rmd"); rmarkdown::render( "vignettes/archival_tags.Rmd", rmarkdown::pdf_document())
```

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
library(walk)

# Additional packages for demo below
library(terra)
library(sf)
library(rnaturalearth)
library(ggplot2)
library(Matrix)
library(ggplot2)
library(tidyterra)
library(ggspatial)
library(lubridate)
library(animation)
```

`walk` estimates a preference function and variation in diffusion rates using a "data likelihood" that represents locational error from tagging or archival tags.  We demonstrate this using data for a tagged Pacific cod in the Aleutian Islands.

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
data(pcod_archival_tag)

# Extract pieces
bathy_terra <- rast(pcod_archival_tag$bathymetry)
likelihood_terra <- rast(pcod_archival_tag$data_likelihood)

# Get land layer
ak <- ne_states( "united states of america", return="sf")
ak <- ak[pmatch("Alas", ak$name_en),]
ak <- st_transform( ak, crs=st_crs("+proj=aea +lat_0=50 +lon_0=-154 +lat_1=55 +lat_2=65 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs") )
ak <- st_geometry(ak)

# change resolution
fact <- 2
bathy_terra <- terra::aggregate( bathy_terra, fact = fact )
likelihood_terra <- as.array( terra::aggregate(likelihood_terra, fact=fact) )

# Eliminate land cells
bathy_terra[bathy_terra<=1] <- NA
bathy_terra$bathy_km <- bathy_terra/1000
bathy_terra$bathy_km2 <- bathy_terra$bathy_km^2
bathy_terra$bathy_t <- terra::ifel(bathy_terra$bathy_km>3, 3, bathy_terra$bathy_km)

# Plot
ggplot() +
  geom_spatraster(data=bathy_terra$bathy_km) + 
  scale_fill_gradientn(colors=rev(sf.colors(10)), na.value = "transparent", name="Bathymetry (km)") + 
  # scale_fill_distiller(palette="YlOrRd", direction = 1, na.value = "transparent", name="UD") + 
  annotation_spatial(ak, fill=gray(0.8), color=1) +
  scale_y_continuous(breaks=seq(-180,180,1)) +
  scale_x_continuous(breaks=seq(-180,180,5)) +
  theme_bw()
```

Next we create data inputs in the format expected:

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
L <- t(apply(likelihood_terra, MARGIN=3, FUN=function(mat){as.vector(t(mat))}))
L <- Matrix(L) |> as("CsparseMatrix")
pdata <- list(L=L)
pdata$times <- data.frame(
  obs=1:nrow(L), 
  timestamp = seq(mdy_hms("2/21/2019 12:00:00"), mdy_hms("5/23/2019 12:00:00"), "1 day")
  )
pdata$times$dt = c(0, diff(pdata$times$timestamp)) 
units(pdata$times$dt) <- "days"
attr(pdata$times,"time_unit") <- "days"

# Format habitat for {walk}
walk_data <- make_walk_data(pdata, bathy_terra, grad=c("bathy_km","bathy_km2"), rast_mask = bathy_terra[[1]]) 

### Bathymetry differences for preference function.
walk_data$q_m$d_bathy <- (walk_data$q_m$bathy_km-walk_data$q_m$from_bathy_km) / (res(bathy_terra)[1]/1000)
walk_data$q_m$d_bathy2 <- (walk_data$q_m$bathy_km^2-walk_data$q_m$from_bathy_km^2) / (res(bathy_terra)[1]/1000)
```

Then we can fit the continuous-time Markov chain as a Hidden Markov model (HMM):

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Diffusion depends on average d_bathy more like traditional CTMC
fit <- fit_ctmc(walk_data,
                 model_parameters = list(
                   q_r = list(form=~ bathy_km), # residency model
                   q_m = list(form=~ d_bathy + d_bathy2) # preference model
                   ), 
                reals = FALSE, fit=TRUE, control=list(trace=1)
                )
```

After model fitting, we can then extract the utilization distribution:

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Get UD
ud <- get_lim_ud(fit)
# Get parameter estimates
beta_q_m <- fit$results$beta$q_m$est
# Calculate preference function from parameters
ud$pref <- beta_q_m[1]*walk_data$q_r$bathy_km + beta_q_m[2]*walk_data$q_r$bathy_km^2
# Get movement-rate (generator) matrix
Q <- get_Q(fit)
# Get expected residence time from diagonal of movement-rate matrix
ud$res <- -1/diag(Q)

# Plot preference as function of bathymetry
plot(walk_data$q_r$bathy_km, ud$pref)
```

Finally, we can plot the 

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Create raster of output
cells <- rast(bathy_terra)
cells[['ud']] <- 0
cells[ud$cell] <- ud$ud
cells[['pref']] <- 0
cells[['pref']][ud$cell] <- ud$pref
cells[['res']] <- 0
cells[['res']][ud$cell] <- ud$res

# Plot utilization distribution
ggplot() +
  geom_spatraster(data=cells$ud) + 
  scale_fill_gradientn(colors=sf.colors(10), na.value = "transparent", name="UD") + 
  # scale_fill_distiller(palette="YlOrRd", direction = 1, na.value = "transparent", name="UD") + 
  annotation_spatial(ak, fill=gray(0.8), color=1) +
  scale_y_continuous(breaks=seq(-180,180,1)) +
  scale_x_continuous(breaks=seq(-180,180,5)) +
  theme_bw()

# Plot preference function
ggplot() +
  geom_spatraster(data=cells$pref) + 
  scale_fill_gradientn(colors=sf.colors(10), na.value = "transparent", name="Preference") + 
  annotation_spatial(ak, fill=gray(0.8), color=1) + 
  scale_y_continuous(breaks=seq(-180,180,1)) +
  scale_x_continuous(breaks=seq(-180,180,5)) +
  theme_bw()

# Plot residence time
ggplot() +
  geom_spatraster(data=cells$res) + 
  scale_fill_gradientn(colors=sf.colors(10), na.value = "transparent", name="E[residence] (d)") + 
  annotation_spatial(ak, fill=gray(0.8), color=1) + 
  scale_y_continuous(breaks=seq(-180,180,1)) +
  scale_x_continuous(breaks=seq(-180,180,5)) +
  theme_bw()
```
